XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-25 talk/icooolps2011/code/map.tex
use listings instead of pygmentize

\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{Map}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps} \PY{o}{=} \PY{p}{\PYZob{}}\PY{p}{\PYZcb{}}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{getindex}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{get}\PY{p}{(}\PY{n}{name}\PY{p}{,} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{)}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{add\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{if} \PY{n}{name} \PY{o+ow}{not} \PY{o+ow}{in} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{:}
            \PY{n}{newmap} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}
            \PY{n}{newmap}\PY{o}{.}\PY{n}{indexes}\PY{o}{.}\PY{n}{update}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{p}{)}
            \PY{n}{newmap}\PY{o}{.}\PY{n}{indexes}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n+nb}{len}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{indexes}\PY{p}{)}
            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]} \PY{o}{=} \PY{n}{newmap}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{other\PYZus{}maps}\PY{p}{[}\PY{n}{name}\PY{p}{]}

\PY{n}{EMPTY\PYZus{}MAP} \PY{o}{=} \PY{n}{Map}\PY{p}{(}\PY{p}{)}

\PY{k}{class} \PY{n+nc}{Instance}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{cls}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls} \PY{o}{=} \PY{n}{cls}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n}{EMPTY\PYZus{}MAP}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage} \PY{o}{=} \PY{p}{[}\PY{p}{]}

    \PY{k}{def} \PY{n+nf}{getfield}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]}
        \PY{k}{raise} \PY{n+ne}{AttributeError}\PY{p}{(}\PY{n}{name}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{write\PYZus{}attribute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{,} \PY{n}{value}\PY{p}{)}\PY{p}{:}
        \PY{n+nb}{map} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n}{index} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{getindex}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{if} \PY{n}{index} \PY{o}{!=} \PY{o}{-}\PY{l+m+mi}{1}\PY{p}{:}
            \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{p}{[}\PY{n}{index}\PY{p}{]} \PY{o}{=} \PY{n}{value}
            \PY{k}{return}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{map} \PY{o}{=} \PY{n+nb}{map}\PY{o}{.}\PY{n}{add\PYZus{}attribute}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{storage}\PY{o}{.}\PY{n}{append}\PY{p}{(}\PY{n}{value}\PY{p}{)}

    \PY{k}{def} \PY{n+nf}{getattr}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{name}\PY{p}{)}\PY{p}{:}
        \PY{k}{try}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{getfield}\PY{p}{(}\PY{n}{name}\PY{p}{)}
        \PY{k}{except} \PY{n+ne}{AttributeError}\PY{p}{:}
            \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{cls}\PY{o}{.}\PY{n}{find\PYZus{}method}\PY{p}{(}\PY{n}{name}\PY{p}{)}
\end{Verbatim}

>>>>

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class Map(object):
    def __init__(self):
        self.indexes = {}
        self.other_maps = {}

    @purefunction
    def getindex(self, name):
        return self.indexes.get(name, -1)

    @purefunction
    def add_attribute(self, name):
        if name not in self.other_maps:
            newmap = Map()
            newmap.indexes.update(self.indexes)
            newmap.indexes[name] = len(self.indexes)
            self.other_maps[name] = newmap
        return self.other_maps[name]

EMPTY_MAP = Map()

class Instance(object):
    def __init__(self, cls):
        self.cls = cls
        self.map = EMPTY_MAP
        self.storage = []

    def getfield(self, name):
        map = hint(self.map, promote=True)
        index = map.getindex(name)
        if index != -1:
            return self.storage[index]
        raise AttributeError(name)

    def write_attribute(self, name, value):
        map = hint(self.map, promote=True)
        index = map.getindex(name)
        if index != -1:
            self.storage[index] = value
            return
        self.map = map.add_attribute(name)
        self.storage.append(value)

    def getattr(self, name):
        try:
            return self.getfield(name)
        except AttributeError:
            return self.cls.find_method(name)
\end{lstlisting}

### 2011-03-25 talk/icooolps2011/paper.tex
use listings instead of pygmentize

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage[scaled=0.8]{beramono}
\usepackage[utf8]{inputenc}

@@

@@

\begin{Verbatim}
inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
\end{Verbatim}

@@

However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{if} \PY{n}{x} \PY{o}{==} \PY{l+m+mi}{4}\PY{p}{:}
    \PY{n}{y} \PY{o}{=} \PY{n}{y} \PY{o}{+} \PY{n}{x}
\end{Verbatim}

If the fragment is traced with \texttt{x} being \texttt{4}, the following trace is
produced:
%
\begin{Verbatim}
guard(x == 4)
y = y + x
\end{Verbatim}

@@

Let's make this more concrete. If we trace a call to the following function:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{def} \PY{n+nf}{f1}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
    \PY{n}{z} \PY{o}{=} \PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
    \PY{k}{return} \PY{n}{z} \PY{o}{+} \PY{n}{y}
\end{Verbatim}

@@

\begin{Verbatim}
v1 = x * 2
z = v1 + 1
v2 = z + y
return(v2)
\end{Verbatim}

Observe how the first two operations could be constant-folded if the value of
\texttt{x} were known. Let's assume that the value of \texttt{x} can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{def} \PY{n+nf}{f2}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
    \PY{n}{x} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n}{x}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
    \PY{n}{z} \PY{o}{=} \PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
    \PY{k}{return} \PY{n}{z} \PY{o}{+} \PY{n}{y}
\end{Verbatim}

@@

\begin{Verbatim}
guard(x == 4)
v1 = x * 2
z = v1 + 1
v2 = z + y
return(v2)
\end{Verbatim}

@@

\begin{Verbatim}
guard(x == 4)
v2 = 9 + y
return(v2)
\end{Verbatim}

@@

\begin{Verbatim}
guard(x == 2)
v2 = 5 + y
return(v2)
\end{Verbatim}

@@

\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{A}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{=} \PY{n}{x}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n}{y}

    \PY{k}{def} \PY{n+nf}{f}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{val}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{compute}\PY{p}{(}\PY{p}{)} \PY{o}{+} \PY{n}{val}

    \PY{k}{def} \PY{n+nf}{compute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
\end{Verbatim}

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
\begin{Verbatim}
x = a.x
v1 = x * 2
v2 = v1 + 1
v3 = v2 + val
a.y = v3
\end{Verbatim}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if \texttt{a} is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is a pure function. To communicate this, there is a \texttt{purefunction} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{Verbatim}[commandchars=\\\{\}]
\PY{k}{class} \PY{n+nc}{A}\PY{p}{(}\PY{n+nb}{object}\PY{p}{)}\PY{p}{:}
    \PY{k}{def} \PY{n+nf}{\PYZus{}\PYZus{}init\PYZus{}\PYZus{}}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{x}\PY{p}{,} \PY{n}{y}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{=} \PY{n}{x}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n}{y}

    \PY{k}{def} \PY{n+nf}{f}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{val}\PY{p}{)}\PY{p}{:}
        \PY{n+nb+bp}{self} \PY{o}{=} \PY{n}{hint}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{,} \PY{n}{promote}\PY{o}{=}\PY{n+nb+bp}{True}\PY{p}{)}
        \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{y} \PY{o}{=} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{compute}\PY{p}{(}\PY{p}{)} \PY{o}{+} \PY{n}{val}

    \PY{n+nd}{@purefunction}
    \PY{k}{def} \PY{n+nf}{compute}\PY{p}{(}\PY{n+nb+bp}{self}\PY{p}{)}\PY{p}{:}
        \PY{k}{return} \PY{n+nb+bp}{self}\PY{o}{.}\PY{n}{x} \PY{o}{*} \PY{l+m+mi}{2} \PY{o}{+} \PY{l+m+mi}{1}
\end{Verbatim}

Now the trace will look like this:
%
\begin{Verbatim}
guard(a == 0xb73984a8)
v1 = compute(a)
v2 = v1 + val
a.y = v2
\end{Verbatim}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{Verbatim}
guard(a == 0xb73984a8)
v2 = 9 + val
a.y = v2
\end{Verbatim}

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage[scaled=0.8]{beramono}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{listings}

@@

\lstset{
  basicstyle=\ttfamily\footnotesize,
  language=Python,
  keywordstyle=\bfseries,
  stringstyle=\color{blue},
  commentstyle=\color{gray}\textit,
  fancyvrb=true,
  showstringspaces=false,
  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
  numberstyle = \tiny,
  numbersep = -20pt,
}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
\end{lstlisting}

@@

However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code:

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
if x == 4:
    y = y + x
\end{lstlisting}

If the fragment is traced with \texttt{x} being \texttt{4}, the following trace is
produced:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(x == 4)
y = y + x
\end{lstlisting}

@@

Let's make this more concrete. If we trace a call to the following function:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f2(x, y):
    x = hint(x, promote=True)
    z = x * 2 + 1
    return z + y
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
v1 = x * 2
z = v1 + 1
v2 = z + y
return(v2)
\end{lstlisting}

Observe how the first two operations could be constant-folded if the value of
\texttt{x} were known. Let's assume that the value of \texttt{x} can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f2(x, y):
    x = hint(x, promote=True)
    z = x * 2 + 1
    return z + y
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(x == 4)
v1 = x * 2
z = v1 + 1
v2 = z + y
return(v2)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(x == 4)
v2 = 9 + y
return(v2)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(x == 2)
v2 = 5 + y
return(v2)
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

    def f(self, val):
        self.y = self.compute() + val

    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
x = a.x
v1 = x * 2
v2 = v1 + 1
v3 = v2 + val
a.y = v3
\end{lstlisting}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if \texttt{a} is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is a pure function. To communicate this, there is a \texttt{purefunction} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

    def f(self, val):
        self = hint(self, promote=True)
        self.y = self.compute() + val

    @purefunction
    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}

Now the trace will look like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(a == 0xb73984a8)
v1 = compute(a)
v2 = v1 + val
a.y = v2
\end{lstlisting}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(a == 0xb73984a8)
v2 = 9 + val
a.y = v2
\end{lstlisting}
